---
title: "Kalman Filter"
author: "Benjamin Panny"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mathematically Defined

$$
\begin{align*}
    \text{Predicted State Estimate} &: \hat{x}_{k|k-1} = F_k \hat{x}_{k-1|k-1} + B_k u_k \\
    \text{Predicted Covariance Estimate} &: P_{k|k-1} = F_k P_{k-1|k-1} F_k^T + Q_k \\
    \text{Innovation or Measurement Residual} &: y_k = z_k - H_k \hat{x}_{k|k-1} \\
    \text{Innovation (Residual) Covariance} &: S_k = H_k P_{k|k-1} H_k^T + R_k \\
    \text{Optimal Kalman Gain} &: K_k = P_{k|k-1} H_k^T S_k^{-1} \\
    \text{Updated State Estimate} &: \hat{x}_{k|k-1} = \hat{x}_{k|k-1} + K_k y_k \\
    \text{Updated Covariance Estimate} &: P_{k|k-1} = (I - K_k H_k) P_{k|k-1}
\end{align*}
$$

$$
\begin{align*}
    &F_k: \text{State Transition model which is applied to the previous state } \hat{x}_{k-1|k-1}. \\
    &B_k: \text{Control Input model which is applied to the control vector } u_k. \\
    &u_k: \text{Control vector at time } k. \\
    &P_{k|k-1}: \text{Predicted Covariance Estimate of the estimation error covariance.} \\
    &Q_k: \text{Process noise covariance matrix, representing the process noise.} \\
    &z_k: \text{Measurement at time } k. \\
    &H_k: \text{Measurement model which maps the true state space into the observed space.} \\
    &y_k: \text{Innovation or Measurement Residual at time } k. \\
    &S_k: \text{Innovation (Residual) Covariance, representing the uncertainty in the measurement.} \\
    &R_k: \text{Measurement noise covariance matrix, representing the measurement noise.} \\
    &K_k: \text{Optimal Kalman Gain, minimizing the a posteriori error covariance.} \\
    &\hat{x}_{k|k}: \text{Updated State Estimate at time } k \text{ given measurement } z_k. \\
    &P_{k|k}: \text{Updated Covariance Estimate of the estimation error covariance.} \\
    &I: \text{Identity matrix, with the same dimensions as } P_{k|k-1}.
\end{align*}
$$


```{r}
# Number of time steps
n <- 100

# True state initialization
true_state <- numeric(n)
true_state[1] <- 0 # initial true state

# Process and measurement noise (variances)
Q <- 1 # process noise variance
R <- 4 # measurement noise variance

# Generate true states and measurements
for (i in 2:n) {
  true_state[i] <- true_state[i-1] + rnorm(1, mean = 0, sd = sqrt(Q))
}
measurements <- true_state + rnorm(n, mean = 0, sd = sqrt(R))

# Kalman Filter initialization
estimated_state <- numeric(n) # To store the estimated states
estimated_state[1] <- 0 # initial estimate
P <- numeric(n) # To store the error covariance
P[1] <- 1 # initial error covariance
K <- numeric(n) # To store the Kalman Gain

for (k in 2:n) {
  # Prediction Update
  predicted_state <- estimated_state[k-1] # since F = 1
  predicted_P <- P[k-1] + Q
  
  # Measurement Update
  K[k] <- predicted_P / (predicted_P + R)
  estimated_state[k] <- predicted_state + K[k] * (measurements[k] - predicted_state)
  P[k] <- (1 - K[k]) * predicted_P
}

# Plot the results
plot(true_state, type = 'l', col = 'blue', ylim = range(c(true_state, measurements, estimated_state)), ylab = 'State', xlab = 'Time')
points(measurements, col = 'red')
lines(estimated_state, col = 'green')
legend('topleft', legend = c('True State', 'Measurements', 'Estimated State'), col = c('blue', 'red', 'green'), lty = 1)
```


This code block does the following:
- It generates a series of true states for a hypothetical system and then simulates measurements of these states with added noise.
- It defines a simple Kalman Filter model using the `KFAS` package, which assumes the process noise and measurement noise covariances are both 1, but you can adjust these values based on your system.
- It applies the Kalman Filter to the measurements to estimate the true states.
- Finally, it plots the true states, the noisy measurements, and the Kalman Filter's estimates of the true states.



