---
title: "3 Variable SCMs"
author: "Benjamin Panny"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

This document contains diagrams, code, and basic math behind 3 variable structural causal models. Including sequential models, mediator models, moderator models, confounder models, and collider models.

## Sequential Models

Sequential models represent a causal sequence where one variable affects another, which in turn affects a third. In the case of three variables, this can be depicted as $A \rightarrow B \rightarrow C$, where $A$ affects $B$, and $B$ affects $C$.

### Diagram

To represent this model graphically, we can use the `DiagrammeR` package in R:

```{r diagram-sequential, echo=TRUE}
library(DiagrammeR)
grViz("
  digraph {
    node [shape = box]
    A -> B -> C
  }
")
```

### Mathematical Formulation

Let's denote:

$X$ as the independent variable,
$M$ as the mediator variable that is affected by $X$ and affects $Y$,
$Y$ as the outcome variable.
The relationships can be described by the following equations:

$M = \alpha + \beta_1X + \epsilon_1$,
$Y = \gamma + \beta_2M + \epsilon_2$,
where $\alpha, \beta_1, \beta_2,$ and $\gamma$ are parameters to be estimated, and $\epsilon_1, \epsilon_2$ are error terms.

```{r}
library(modelsummary)
set.seed(123) # For reproducibility
n <- 100 # Number of observations

# Simulating the independent variable X
X <- rnorm(n, mean = 0, sd = 1)

# Simulating the mediator variable M
M <- 0.5 + 0.4*X + rnorm(n, mean = 0, sd = 1)

# Simulating the outcome variable Y
Y <- -0.2 + 0.5*M + rnorm(n, mean = 0, sd = 1)

# Fitting the model
seq_model <- list(
  "x_and_m" <- lm(Y ~ X + M),
  "x" <- lm (Y ~ X),
  "m" <- lm(Y ~ M)
)
modelsummary(seq_model, output = 'kableExtra')
```

