---
title: "Predator-Prey Simulation with Lotka-Volterra Equations"
author: "Benjamin Panny"
date: "`r Sys.Date()`"
output: github_document
---
```{r}

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lotka-Volterra Equations

The Lotka-Volterra equations, used to simulate predator-prey interactions, are a pair of first-order, non-linear, differential equations. They describe the dynamics of biological systems where two species interact, one as a predator and the other as prey:

\[
\frac{dP}{dt} = \alpha P - \beta P H
\]

\[
\frac{dH}{dt} = -\gamma H + \delta P H
\]

where:

- \( P \) is the number of preys (e.g., rabbits),
- \( H \) is the number of predators (e.g., foxes),
- \( \frac{dP}{dt} \) and \( \frac{dH}{dt} \) represent the growth rates of prey and predator populations,
- \( \alpha, \beta, \gamma, \) and \( \delta \) are positive real constants that describe the interaction of the two species.

The phase space plot shows the trajectories of both species in a graph, with the number of prey as the x-axis and the number of predators as the y-axis.


```{r}
library(deSolve)
library(ggplot2)

# Lotka-Volterra Model
lotka_volterra <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dP <- alpha * P - beta * P * H
    dH <- -gamma * H + delta * P * H
    list(c(dP, dH))
  })
}

# Parameters
params <- c(alpha = 0.1,  # Prey birth rate
            beta = 0.02,  # Prey death rate per predator
            gamma = 0.3,  # Predator death rate
            delta = 0.01) # Predator birth rate per prey

# Initial State
state <- c(P = 40,  # Initial prey population
           H = 9)   # Initial predator population

# Time
times <- seq(0, 200, by = 0.1)

# Solving the differential equations
out <- ode(y = state, times = times, func = lotka_volterra, parms = params)

# Creating a data frame for plotting
out_df <- as.data.frame(out)
colnames(out_df)[2:3] <- c("Prey", "Predators")

# Plotting the phase space
phase_space_plot <- ggplot(out_df, aes(x = Prey, y = Predators)) +
  geom_path(color = "blue") +
  theme_minimal() +
  labs(title = "Predator-Prey Phase Space",
       x = "Number of Prey",
       y = "Number of Predators")

print(phase_space_plot)
```

# Simulate Over Parameter Grid

```{r}
# Extended parameter grid
params_grid <- expand.grid(alpha = seq(0.1, 0.3, length.out = 3),
                           beta = seq(0.02, 0.05, length.out = 3),
                           gamma = seq(0.3, 0.5, length.out = 3),
                           delta = seq(0.01, 0.03, length.out = 3),
                           P = seq(30, 50, length.out = 3),
                           H = seq(5, 15, length.out = 3))

# Time
times <- seq(0, 100, by = 0.1)

# Run simulations for each parameter set
all_sims <- do.call(rbind, lapply(1:nrow(params_grid), function(i) {
  params <- as.list(params_grid[i, ])
  state <- c(P = params$P, H = params$H)
  out <- ode(y = state, times = times, func = lotka_volterra, parms = params)
  out_df <- as.data.frame(out)
  out_df$params <- paste(params$alpha, params$beta, params$gamma, params$delta, params$P, params$H, sep = "_")
  # Or instead assign each parameter to its own column in the data frame
  # out_df$alpha <- params$alpha
  # out_df$beta <- params$beta
  # out_df$gamma <- params$gamma
  # out_df$delta <- params$delta
  # out_df$P_init <- params$P
  # out_df$H_init <- params$H
  out_df
}))

# Or you can separate the parameters out separately
# all_sims %>% 
#   separate(params, into = c("alpha", "beta", "gamma", "delta", "P_init", "H_init"), sep = "_")
```

# Animate the Results Over the Parameter Grid

Doing it all at once is a lot, and a little confusing to change all the parameters at once, so let's do it one at a time!

```{r}
library(gganimate)
# 
# Create the ggplot object
# p <- ggplot(all_sims, aes(x = P, y = H, group = params)) +
#   geom_path(aes(color = params)) +
#   labs(x = "Number of Prey", y = "Number of Predators") +
#   theme_minimal()
# 
# Animate the plot with gganimate
# animation <- p + 
#   transition_states(params, transition_length = 2, state_length = 1) +
#   enter_fade() + exit_fade()
# 
# 
# Render and save the animation
# anim_save("lotka_volterra_extended_animation.gif", animate(animation, nframes = 100, fps = 10))
```

```{r}
library(dplyr)

split_sims <- all_sims %>% 
  separate(params, into = c("alpha", "beta", "gamma", "delta", "P_init", "H_init"), sep = "_") %>% 
  mutate(across(where(is.character), ~as.numeric(.))) 

# Fixed parameter values
fixed_alpha <- 0.1
fixed_beta <- 0.02
fixed_gamma <- 0.3
fixed_delta <- 0.01
fixed_H_init <- 10

# Filter all_sims for simulations with only the fixed parameters and varying P_init
filtered_sims <- split_sims %>%
  filter(alpha == fixed_alpha, 
         beta == fixed_beta, 
         gamma == fixed_gamma, 
         delta == fixed_delta, 
         H_init == fixed_H_init)
```

```{r}
library(ggplot2)
library(gganimate)

# Assuming filtered_sims is your data frame with simulations for varying P_init

# Create the ggplot object
p <- ggplot(filtered_sims, aes(x = P, y = H, group = P_init)) +
  geom_path(aes(color = as.factor(P_init))) +
  labs(title = 'Predator-Prey Model with Varying Initial Prey Population',
       subtitle = 'Initial Prey Population: {closest_state}',
       x = "Number of Prey", 
       y = "Number of Predators") +
  theme_minimal() # +
  # theme(legend.position = "none") # Remove the legend if not needed

# Animate the plot with gganimate
animation <- p + 
  transition_states(P_init, transition_length = 2, state_length = 1) # +
  # enter_fade() + exit_fade()

# animate(animation, nframes = 100, fps = 10)

# Render and save the animation
anim_save("lotka_volterra_varying_p_init.gif", animate(animation, nframes = 100, fps = 10))
```




