---
title: "Many Mathematical Functions"
author: "Benjamin Panny"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Algebraic Functions

Algebraic functions are functions that can be expressed as the solution of a polynomial equation with integer coefficients.

## Polynomials: Can be generated solely by addition, multiplication, and raising to the power of a positive integer.

As can be seen the below plots, the highest degree determines the characteristic shape of the plot!

### Constant function: polynomial of degree zero, graph is a horizontal straight line

```{r}
ggplot(data.frame(x = c(-10, 10)), aes(x)) +
  geom_line(aes(y = 5), color = "blue") +
  ggtitle("Constant Function y=5") +
  theme_minimal()
```


### Linear function: First degree polynomial, graph is a straight line.

```{r}
ggplot(data.frame(x = c(-10, 10)), aes(x)) +
  geom_line(aes(y = 2*x + 3), color = "green") +
  ggtitle("Linear Function y=2x+3") +
  theme_minimal()
```


### Quadratic function: Second degree polynomial, graph is a parabola.

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = x^2 - 4*x + 4), color = "red") +
  ggtitle("Quadratic Function y=x^2-4x+4") +
  theme_minimal()
```


### Cubic function: Third degree polynomial.

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = x^3 - 6*x^2 + 11*x - 6), color = "purple") +
  ggtitle("Cubic Function y=x^3-6x^2+11x-6") +
  theme_minimal()
```


### Quartic function: Fourth degree polynomial.

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = x^4 - 10*x^3 + 35*x^2 - 50*x + 24), color = "orange") +
  ggtitle("Quartic Function y=x^4-10x^3+35x^2-50x+24") +
  theme_minimal()
```


### Quintic function: Fifth degree polynomial.

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = x^5 - 10*x^4 + 40*x^3 - 80*x^2 + 80*x - 32), color = "brown") +
  ggtitle("Quintic Function y=x^5-10x^4+40x^3-80x^2+80x-32") +
  theme_minimal()
```


### Sextic function: Sixth degree polynomial.

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = x^6 - 6*x^5 + 15*x^4 - 20*x^3 + 15*x^2 - 6*x + 1), color = "cyan") +
  ggtitle("Sextic Function y=x^6-6x^5+15x^4-20x^3+15x^2-6x+1") +
  theme_minimal()
```


### Rational functions: A ratio of two polynomials.

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = (x^2 + 2*x + 2) / (x^2 - 4)), color = "magenta") +
  ggtitle("Rational Function y=(x^2+2x+2)/(x^2-4)") +
  theme_minimal() +
  ylim(-50, 50)
```


## nth root

### Square root: Yields a number whose square is the given one.

```{r}
ggplot(data.frame(x = seq(0, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = sqrt(x)), color = "black") +
  ggtitle("Square Root Function y=sqrt(x)") +
  theme_minimal()
```


### Cube root: Yields a number whose cube is the given one.

```{r}
ggplot(data.frame(x = seq(-100, 100, length.out = 400)), aes(x = x)) +
  geom_line(aes(y = sign(x) * abs(x)^(1/3)), color = "darkgreen") +
  ggtitle("Cube Root Function y=cube root(x)") +
  theme_minimal()
```

### Quartic root: Yields a number whose square is the given one.

```{r}
ggplot(data.frame(x = seq(0, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = x^(1/4)), color = "black") +
  ggtitle("Quartic Root Function y=sqrt(x)") +
  theme_minimal()
```


### Quintic root: Yields a number whose cube is the given one.

```{r}
ggplot(data.frame(x = seq(-100, 100, length.out = 400)), aes(x = x)) +
  geom_line(aes(y = sign(x) * abs(x)^(1/5)), color = "darkgreen") +
  ggtitle("Quintic Root Function y=cube root(x)") +
  theme_minimal()
```

# Transcendental functions are functions that are not algebraic (non-polynomial)

## Exponential function: raises a fixed number to a variable power.

```{r}
ggplot(data.frame(x = seq(-2, 2, length.out = 400)), aes(x)) +
  geom_line(aes(y = exp(x)), color = "blue") +
  ggtitle("Exponential Function y=exp(x)") +
  theme_minimal()
```

## Hyperbolic functions: formally similar to the trigonometric functions.

### Sine Hyperbolic

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = sinh(x)), color = "blue") +
  ggtitle("Hyperbolic Sine Function y=sinh(x)") +
  theme_minimal()
```

### Cosine Hyperbolic

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = cosh(x)), color = "red") +
  ggtitle("Hyperbolic Cosine Function y=cosh(x)") +
  theme_minimal()

```

### Tangent Hyperbolic

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = tanh(x)), color = "green") +
  ggtitle("Hyperbolic Tangent Function y=tanh(x)") +
  theme_minimal()
```



## Inverse hyperbolic functions: inverses of the hyperbolic functions, analogous to the inverse circular functions.

### Sine Hyperbolic

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = asinh(x)), color = "blue") +
  ggtitle("Inverse Hyperbolic Sine Function y=asinh(x)") +
  theme_minimal()
```

### Cosine Hyperbolic

```{r}
ggplot(data.frame(x = seq(-5, 50, length.out = 400)), aes(x)) +
  geom_line(aes(y = acosh(x)), color = "red") +
  ggtitle("Inverse Hyperbolic Cosine Function y=acosh(x)") +
  theme_minimal()
```

### Inverse Tangent Hyperbolic

```{r}
ggplot(data.frame(x = seq(-3, 3, length.out = 400)), aes(x)) +
  geom_line(aes(y = atanh(x)), color = "green") +
  ggtitle("Inverse Hyperbolic Tangent Function y=tanh(x)") +
  theme_minimal()
```

## Logarithms: the inverses of exponential functions; useful to solve equations involving exponentials.

### Natural logarithm

```{r}
ggplot(data.frame(x = seq(0.01, 5, length.out = 400)), aes(x)) +
  geom_line(aes(y = log(x)), color = "red") +
  ggtitle("Natural Logarithm y=log(x)") +
  theme_minimal()
```

### Common logarithm

```{r}
ggplot(data.frame(x = seq(0.01, 5, length.out = 400)), aes(x)) +
  geom_line(aes(y = log10(x)), color = "green") +
  ggtitle("Common Logarithm y=log10(x)") +
  theme_minimal()
```

### Binary logarithm

```{r}
ggplot(data.frame(x = seq(0.01, 5, length.out = 400)), aes(x)) +
  geom_line(aes(y = log2(x)), color = "green") +
  ggtitle("Common Logarithm y=log10(x)") +
  theme_minimal()
```

## Power functions: raise a variable number to a fixed power; also known as Allometric functions; note: if the power is a rational number it is not strictly a transcendental function.


```{r}
ggplot(data.frame(x = seq(-5, 5, length.out = 400)), aes(x)) +
  geom_line(aes(y = x^3), color = "green") +
  ggtitle("y=x") +
  theme_minimal()
```

## Periodic (Trignometric) functions: 

c

### Cotangent

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1/tan(x)), color = "blue") +
  ggtitle("Cotangent Function y=cot(x)") +
  theme_minimal() +
  ylim(-10, 10)

```

### Secant

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1/cos(x)), color = "red") +
  ggtitle("Secant Function y=sec(x)") +
  theme_minimal() +
  ylim(-10, 10)

```

### Cosecant

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1/sin(x)), color = "green") +
  ggtitle("Cosecant Function y=csc(x)") +
  theme_minimal() +
  ylim(-10, 10)

```

### Exsecant

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1/cos(x) - 1), color = "purple") +
  ggtitle("Exsecant Function y=exsec(x)") +
  theme_minimal() +
  ylim(-10, 10)

```

### Excosecant

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1/sin(x) - 1), color = "orange") +
  ggtitle("Excosecant Function y=excsc(x)") +
  theme_minimal() +
  ylim(-10, 10)

```

### Versine

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1 - cos(x)), color = "cyan") +
  ggtitle("Versine Function y=versin(x)") +
  theme_minimal()

```

### Coversine 

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1 - sin(x)), color = "magenta") +
  ggtitle("Coversine Function y=coversin(x)") +
  theme_minimal()

```

### Vercosine 

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1 + cos(x)), color = "yellow") +
  ggtitle("Vercosine Function y=vercosin(x)") +
  theme_minimal()

```

### Covercosine 

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = 1 + sin(x)), color = "black") +
  ggtitle("Covercosine Function y=covercosin(x)") +
  theme_minimal()

```

### Haversine 

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = (1 - cos(x))/2), color = "blue") +
  ggtitle("Haversine Function y=haversin(x)") +
  theme_minimal()

```

### Hacoversine 

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = (1 - sin(x))/2), color = "red") +
  ggtitle("Hacoversine Function y=hacoversin(x)") +
  theme_minimal()

```

### Havercosine 

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = (1 + cos(x))/2), color = "green") +
  ggtitle("Havercosine Function y=havercosin(x)") +
  theme_minimal()

```

### Hacovercosine 

```{r}
ggplot(data.frame(x = seq(-pi, 3*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = (1 + sin(x))/2), color = "purple") +
  ggtitle("Hacovercosine Function y=hacovercosin(x)") +
  theme_minimal()

```

## Inverse trigonometric functions etc.; used in geometry and to describe periodic phenomena. 

### Inverse Sine (arcsin)

```{r}
ggplot(data.frame(x = seq(-2*pi, 2*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = asin(x)), color = "purple") +
  ggtitle("Inverse Sine Function y=arcsin(x)") +
  theme_minimal()
```

### Inverse Cosine (arccos) 

```{r}
ggplot(data.frame(x = seq(-2*pi, 2*pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = acos(x)), color = "orange") +
  ggtitle("Inverse Cosine Function y=arccos(x)") +
  theme_minimal()
```

### Inverse Tangent (arctan)

```{r}
ggplot(data.frame(x = seq(-pi, pi, length.out = 400)), aes(x)) +
  geom_line(aes(y = atan(x)), color = "green") +
  ggtitle("Inverse Tangent Function y=arctan(x)") +
  theme_minimal() +
  ylim(-10, 10)  # Limit y-axis to avoid extreme values near pi/2 and -pi/2
```

### Gudermannian function.

The Gudermannian function relates the circular functions and the hyperbolic ones without involving complex numbers.

```{r}
ggplot(data.frame(x = seq(-10, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = 2*atan(exp(x)) - pi/2), color = "darkgreen") +
  ggtitle("Gudermannian Function") +
  theme_minimal()
```

# Special functions
## Piecewise special functions
### Indicator function: maps x to either 1 or 0, depending on whether or not x belongs to some subset. 
The indicator function can be defined for a subset, such as x > 0.

```{r}
ggplot(data.frame(x = seq(-2, 2, length.out = 400)), aes(x)) +
  geom_point(aes(y = ifelse(x > 0, 1, 0)), color = "blue") +
  ggtitle("Indicator Function") +
  theme_minimal()
```

### Step function: A finite linear combination of indicator  functions of half-open intervals.

```{r}
ggplot(data.frame(x = seq(-2, 2, length.out = 100)), aes(x)) +
  geom_point(aes(y = ifelse(x > 0, 1, ifelse(x < -1, -1, 0))), color = "blue") +
  ggtitle("Step Function") +
  theme_minimal()
```

#### Heaviside step function: 0 for negative arguments and 1 for positive arguments. The integral of the Dirac delta function.

```{r}
ggplot(data.frame(x = seq(-2, 2, length.out = 400)), aes(x)) +
  geom_point(aes(y = ifelse(x >= 0, 1, 0)), color = "red") +
  ggtitle("Heaviside Step Function") +
  theme_minimal()
```

### Sawtooth wave

The sawtooth wave can be created using the modulo operation to repeat the pattern.

```{r}
ggplot() +
  geom_function(fun = function(x) x - floor(x), aes(color = "green")) +
  xlim(0, 10) +
  ggtitle("Sawtooth Wave") +
  theme_minimal()
```

### Square wave

A square wave can alternate between -1 and 1. We use the modulo operation and ifelse for simplicity.

```{r}
ggplot(data.frame(x = seq(0, 10, length.out = 400)), aes(x)) +
  geom_line(aes(y = ifelse(floor(x) %% 2 == 0, 1, -1)), color = "purple") +
  ggtitle("Square Wave") +
  theme_minimal()
```

### Triangle wave

Similar to a sawtooth but with an absolute value to create a triangle shape.

```{r}

ggplot() +
  geom_function(fun = function(x)  abs(1 - 2 * ((x - 0.5) %% 1)), color = "green") +
  xlim(-5, 5) +
  ggtitle("Triangle Wave") +
  theme_minimal()
```

### Rectangular function

```{r}
ggplot(data.frame(x = seq(-2, 2, length.out = 400)), aes(x)) +
  geom_line(aes(y = ifelse(x >= -1 & x <= 1, 1, 0)), color = "blue") +
  ggtitle("Rectangular Function") +
  theme_minimal()
```

### Floor function: Largest integer less than or equal to a given number.

```{r}
ggplot(data.frame(x = seq(-3, 3, length.out = 400)), aes(x)) +
  geom_point(aes(y = floor(x)), color = "magenta") +
  ggtitle("Floor Function") +
  theme_minimal()
```

### Ceiling function: Smallest integer larger than or equal to a given number.

```{r}
ggplot(data.frame(x = seq(-3, 3, length.out = 400)), aes(x)) +
  geom_point(aes(y = ceiling(x)), color = "magenta") +
  theme_minimal()
  ggtitle("Ceiling Function") +
```

### Sign function: Returns only the sign of a number, as +1 or −1.

```{r}
ggplot() +
  geom_function(fun = function(x) sign(x), color = "blue") +
  xlim(-10, 10) +
  ggtitle("Sign Function") +
  theme_minimal()
```

### Absolute value: distance to the origin (zero point)

```{r}
ggplot() +
  geom_function(fun = function(x) abs(x), color = "red") +
  xlim(-10, 10) +
  ggtitle("Absolute Value") +
  theme_minimal()
```

## Arithmetic functions

### Sigma function: Sums of powers of divisors of a given natural number.

The Sigma function, $\sigma_k(n)$ sums the $k^th$ powers of the divisors of $n$. Here is $\sigma_2(n)$, the squared sum of divisors

```{r}
sigma_function <- function(n) {
  divisors <- seq(1, n)
  sum(divisors[n %% divisors == 0])
}

ggplot(data.frame(x = seq(1, 1000, length.out = 1000)), aes(x)) +
  geom_point(aes(y = sapply(x, sigma_function)), color = "magenta") +
  theme_minimal() +
  ggtitle("Sigma Function")
```


### Euler's totient function: Number of numbers coprime to (and not bigger than) a given one.

```{r}
euler_totient <- function(n) {
  if (n == 1) return(1)
  length(Filter(function(x) gcd(x, n) == 1, 1:n))
}

# Example: Compute Euler's totient function for numbers 1 to 10
sapply(1:10, euler_totient)
```

### Prime-counting function: Number of primes less than or equal to a given number.
### Partition function: Order-independent count of ways to write a given positive integer as a sum of positive integers.
### Möbius μ function: Sum of the nth primitive roots of unity, it depends on the prime factorization of n.
### Prime omega functions
### Chebyshev functions
### Liouville function, λ(n) = (–1)Ω(n)
### Von Mangoldt function, Λ(n) = log p if n is a positive power of the prime p
### Carmichael function

## Antiderivatives of elementary functions
### Logarithmic integral function: Integral of the reciprocal of ## the logarithm, important in the prime number theorem.
### Exponential integral
### Trigonometric integral: Including Sine Integral and Cosine ## Integral
### Inverse tangent integral
### Error function: An integral important for normal random variables.
#### Fresnel integral: related to the error function; used in optics.
#### Dawson function: occurs in probability.
#### Faddeeva function
## Gamma and related functions
## Gamma function: A generalization of the factorial function.
## Barnes G-function
## Beta function: Corresponding binomial coefficient analogue.
## Digamma function, Polygamma function
## Incomplete beta function
## Incomplete gamma function
## K-function
## Multivariate gamma function: A generalization of the Gamma ## function useful in multivariate statistics.
## Student's t-distribution
## Pi function 
## Elliptic and related functions
## Elliptic integrals: Arising from the path length of ellipses; ## important in many applications. Alternate notations include:
## Carlson symmetric form
## Legendre form
## Nome
## Quarter period
## Elliptic functions: The inverses of elliptic integrals; used ## to model double-periodic phenomena.
## Jacobi's elliptic functions
## Weierstrass's elliptic functions
## Lemniscate elliptic functions
## Theta functions
## Neville theta functions
## Modular lambda function
## Closely related are the modular forms, which include
## J-invariant
## Dedekind eta function
## Bessel and related functions
## Airy function
## Bessel functions: Defined by a differential equation; useful ## in astronomy, electromagnetism, and mechanics.
## Bessel–Clifford function
## Kelvin functions
## Legendre function: From the theory of spherical harmonics.
## Scorer's function
## Sinc function
## Hermite polynomials
## Laguerre polynomials
## Chebyshev polynomials
## Synchrotron function
## Riemann zeta and related functions
## Riemann zeta function: A special case of Dirichlet series.
## Riemann Xi function
## Dirichlet eta function: An allied function.
## Dirichlet beta function
## Dirichlet L-function
## Hurwitz zeta function
## Legendre chi function
## Lerch transcendent
## Polylogarithm and related functions:
## Incomplete polylogarithm
## Clausen function
## Complete Fermi–Dirac integral, an alternate form of the ## polylogarithm.
## Dilogarithm
## Incomplete Fermi–Dirac integral
## Kummer's function
## Riesz function
## Hypergeometric and related functions
## Hypergeometric functions: Versatile family of power series.
## Confluent hypergeometric function
## Associated Legendre functions
## Meijer G-function
## Fox H-function
## Iterated exponential and related functions
## Hyper operators
## Iterated logarithm
## Pentation
## Super-logarithms
## Super-roots
## Tetration
## Other standard special functions
## Lambert W function: Inverse of f(w) = w exp(w).
## Lamé function
## Mathieu function
## Mittag-Leffler function
## Painlevé transcendents
## Parabolic cylinder function
## Arithmetic–geometric mean
## Miscellaneous functions
## Ackermann function: in the theory of computation, a ## computable function that is not primitive recursive.
## Dirac delta function: everywhere zero except for x = 0; total ## integral is 1. Not a function but a distribution, but ## sometimes informally referred to as a function, particularly ## by physicists and engineers.
## Dirichlet function: is an indicator function that matches 1 ## to rational numbers and 0 to irrationals. It is nowhere ## continuous.
## Thomae's function: is a function that is continuous at all ## irrational numbers and discontinuous at all rational numbers. ## It is also a modification of Dirichlet function and sometimes ## called Riemann function.
## Kronecker delta function: is a function of two variables, ## usually integers, which is 1 if they are equal, and 0 ## otherwise.
## Minkowski's question mark function: Derivatives vanish on the ## rationals.
## Weierstrass function: is an example of continuous function ## that is nowhere differentiable
