---
title: "basics"
author: "Benjamin Panny"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Matrix Addition

Matrix addition is performed element-wise between two matrices of the same dimensions. If $A$ and $B$ are two matrices of size $m \times n$, their sum $C$ is also an $m \times n$ matrix where each element $c_{ij} = a_{ij} + b_{ij}$.

$$
C = A + B
$$
For example, if:

$$
A = \begin{pmatrix}
1 & 2 \\
3 & 4 \\
\end{pmatrix}, \quad
B = \begin{pmatrix}
2 & 3 \\
4 & 5 \\
\end{pmatrix}
$$

Then

$$
C = \begin{pmatrix}
1+2 & 2+3 \\
3+4 & 4+5 \\
\end{pmatrix} = \begin{pmatrix}
3 & 5 \\
7 & 9 \\
\end{pmatrix}
$$

```{r}
A <- matrix(c(1, 2, 3, 4), byrow = T, nrow = 2)
B <- matrix(c(2, 3, 4, 5), byrow = T, nrow = 2)

# Matrix Addition
C <- A + B
print(C)
```

# Matrix Multiplication
Matrix multiplication involves the dot product of rows of the first matrix with columns of the second matrix. For two matrices $A$ of size $m \times n$ and $B$ of size $n \times p$, the product $AB$ will be of size $m \times p$ where each element $c_{ij} = \sum_{k=1}^{n} a_{ik} b_{kj}$.

$$
AB = C
$$

For example, if

$$
A = \begin{pmatrix}
1 & 2 \\
3 & 4 \\
\end{pmatrix}, \quad
B = \begin{pmatrix}
2 & 0 \\
1 & 2 \\
\end{pmatrix}
$$

Then

$$
C = \begin{pmatrix}
(1)(2)+(2)(1) & (1)(0)+(2)(2) \\
(3)(2)+(4)(1) & (3)(0)+(4)(2) \\
\end{pmatrix} = \begin{pmatrix}
4 & 4 \\
10 & 8 \\
\end{pmatrix}
$$

```{r}
A <- matrix(c(1,3,2,4), nrow=2)
B <- matrix(c(2,1,0,2), nrow=2)
C <- A %*% B
print(C)
```

# Determinant

The determinant of a matrix is a scalar value that can be computed from the elements of a square matrix and encodes certain properties of the matrix. For a 2x2 matrix:

$$
A = \begin{pmatrix}
a & b \\
c & d \\
\end{pmatrix}
$$

The determinant of A is:

$$
\text{det}(A) = ad - bc 
$$

The determinant is important for understanding matrix invertibility; a matrix is invertible if and only if its determinant is non-zero.

#Matrix Inversion

Matrix inversion can be achieved through the method of minors and cofactors, as well as by Gauss-Jordan elimination. For a matrix $A$, its inverse $A^{-1}$ satisfies the equation $AA^{-1} = I$, where $I$ is the identity matrix.

$$
A^{-1} = \frac{1}{\text{det}(A)} \begin{pmatrix}
d & -b \\
-c & a \\
\end{pmatrix}
$$

# Linear System of Equations

A system of linear equations can be represented in matrix form as $Ax=b$, where 
$A$ is a matrix of coefficients, $x$ is a column vector of variables, and $b$ is a column vector of results.

$$
Ax = b
$$

For example

$$
\begin{align*}
x + 2y &= 5 \\
3x + 4y &= 11
\end{align*}
$$

```{r}
eq1 <- function(x) { (5 - x) / 2 }
eq2 <- function(x) { (11 - 3*x) / 4 }

x_vals <- seq(-10, 10, length.out = 200)
y_vals1 <- eq1(x_vals)
y_vals2 <- eq2(x_vals)

plot(x_vals, y_vals1, type='l', col='blue', ylim=c(-10, 10), ylab='y', xlab='x')
lines(x_vals, y_vals2, col='red')
legend("topright", legend=c("x + 2y = 5", "3x + 4y = 11"), col=c("blue", "red"), lty=1)
```

```{r}
A <- matrix(c(1, 3, 2, 4), nrow=2)
b <- c(5, 11)

solution <- solve(A, b)

plot(x_vals, y_vals1, type='l', col='blue', ylim=c(-10, 10), ylab='y', xlab='x')
lines(x_vals, y_vals2, col='red')
points(solution[1], solution[2], pch=19, col='green')
legend("topright", legend=c("x + 2y = 5", "3x + 4y = 11", "Solution"), col=c("blue", "red", "green"), lty=1, pch=c(NA, NA, 19))
```

# Vector Combinations

Vector Space and Linear Combinations
A vector space is a collection of vectors that can be scaled and added together. A linear combination of a set of vectors ${v_1, v_2, \ldots, v_n}$ involves adding together these vectors, each multiplied by a scalar. This can be expressed as:

$$
c_1v_1 + c_2v_2 + \ldots + c_nv_n
$$

$c_1, c_2, \ldots c_n$ are scalars.

For example, given vectors $v_1 = (2,3), \quad v_2 = (-1,2)$, find $v$ that is a linear combination of $v1$ and $v2$ such that $v = (0,8)$

$$
v = c_1v_1 + c_2v_2
$$

We need  t o solve for $c_1$ and $c_2$

$$
c_1(2, 3) + c_2(-1, 2) = (0, 8)
$$

Expanding the equation, we get:

$$
(2c_1 - c_2, 3c_1 + 2c_2) = (0, 8)
$$

Which is the system of equations:

$$
\begin{pmatrix}
2  & -1  \\
3 & 2 \\
\end{pmatrix}
\begin{pmatrix}
c_1  \\
c_2 \\
\end{pmatrix}
= 
\begin{pmatrix}
0  \\
8 \\
\end{pmatrix}
$$

```{r}
A <- matrix(c(2, 3, -1, 2), nrow=2)
b <- c(0, 8)

solution <- solve(A, b)
cat("c_1 =", solution[1], "c_2 =", solution[2], "\n")
```

To visualize the vectors \(v_1\), \(v_2\), and their linear combination \(v\), we use `ggplot2`. This visualization helps in understanding how the combination of \(v_1\) and \(v_2\) results in \(v\).

```{r plot-vectors, fig.cap="Visualization of Vectors and Their Linear Combination"}
library(ggplot2)

# Define the original vectors and their combination
v1 <- c(2, 3)
v2 <- c(-1, 2)
v <- solution[1] * v1 + solution[2] * v2

# Create a data frame for plotting
vectors_df <- data.frame(
  x = c(0, 0, 0),
  y = c(0, 0, 0),
  xend = c(v1[1], v2[1], v[1]),
  yend = c(v1[2], v2[2], v[2]),
  color = c("v1", "v2", "v")
)

# Plot
ggplot() +
  geom_segment(data = vectors_df, aes(x = x, y = y, xend = xend, yend = yend, color = color), arrow = arrow(type = "closed")) +
  scale_color_manual(values = c("v1" = "blue", "v2" = "red", "v" = "green")) +
  labs(title = "Original Vectors", x = "X", y = "Y") +
  theme_minimal()

# Create a data frame for plotting
vectors_df <- data.frame(
  x = c(0, v1[1]*solution[1], 0),
  y = c(0, v1[2]*solution[1], 0),
  xend = c(v1[1]*solution[1], v1[1]*solution[1]+v2[1]*solution[2], v[1]),
  yend = c(v1[2]*solution[1], v1[2]*solution[1]+v2[2]*solution[2], v[2]),
  color = c("v1", "v2", "v")
)

ggplot() +
  geom_segment(data = vectors_df, aes(x = x, y = y, xend = xend, yend = yend, color = color), arrow = arrow(type = "closed")) +
  scale_color_manual(values = c("v1" = "blue", "v2" = "red", "v" = "green")) +
  labs(title = "Combined Vectors", x = "X", y = "Y") +
  theme_minimal()
```

This example demonstrates solving a vector combination problem by finding scalar multipliers that combine given vectors to result in a specific vector. The R code solves for the scalars and visualizes the vectors and their combination.

Applications of vector combinations include solving systems of linear equations, transforming geometrical shapes in computer graphics, and modeling physical systems in engineering and physics. Linear combinations are also foundational in understanding vector spaces, subspaces, and spans in linear algebra.
