---
title: "basics"
author: "Benjamin Panny"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Matrix Addition

Matrix addition is performed element-wise between two matrices of the same dimensions. If $A$ and $B$ are two matrices of size $m \times n$, their sum $C$ is also an $m \times n$ matrix where each element $c_{ij} = a_{ij} + b_{ij}$.

$$
C = A + B
$$
For example, if:

$$
A = \begin{pmatrix}
1 & 2 \\
3 & 4 \\
\end{pmatrix}, \quad
B = \begin{pmatrix}
2 & 3 \\
4 & 5 \\
\end{pmatrix}
$$

Then

$$
C = \begin{pmatrix}
1+2 & 2+3 \\
3+4 & 4+5 \\
\end{pmatrix} = \begin{pmatrix}
3 & 5 \\
7 & 9 \\
\end{pmatrix}
$$

```{r}
A <- matrix(c(1, 2, 3, 4), byrow = T, nrow = 2)
B <- matrix(c(2, 3, 4, 5), byrow = T, nrow = 2)

# Matrix Addition
C <- A + B
print(C)
```

# Matrix Multiplication
Matrix multiplication involves the dot product of rows of the first matrix with columns of the second matrix. For two matrices $A$ of size $m \times n$ and $B$ of size $n \times p$, the product $AB$ will be of size $m \times p$ where each element $c_{ij} = \sum_{k=1}^{n} a_{ik} b_{kj}$.

$$
AB = C
$$

For example, if

$$
A = \begin{pmatrix}
1 & 2 \\
3 & 4 \\
\end{pmatrix}, \quad
B = \begin{pmatrix}
2 & 0 \\
1 & 2 \\
\end{pmatrix}
$$

Then

$$
C = \begin{pmatrix}
(1)(2)+(2)(1) & (1)(0)+(2)(2) \\
(3)(2)+(4)(1) & (3)(0)+(4)(2) \\
\end{pmatrix} = \begin{pmatrix}
4 & 4 \\
10 & 8 \\
\end{pmatrix}
$$

```{r}
A <- matrix(c(1,3,2,4), nrow=2)
B <- matrix(c(2,1,0,2), nrow=2)
C <- A %*% B
print(C)
```

# Determinant

The determinant of a matrix is a scalar value that can be computed from the elements of a square matrix and encodes certain properties of the matrix. For a 2x2 matrix:

$$
A = \begin{pmatrix}
a & b \\
c & d \\
\end{pmatrix}
$$

The determinant of A is:

$$
\text{det}(A) = ad - bc 
$$

The determinant is important for understanding matrix invertibility; a matrix is invertible if and only if its determinant is non-zero.

#Matrix Inversion

Matrix inversion can be achieved through the method of minors and cofactors, as well as by Gauss-Jordan elimination. For a matrix $A$, its inverse $A^{-1}$ satisfies the equation $AA^{-1} = I$, where $I$ is the identity matrix.

$$
A^{-1} = \frac{1}{\text{det}(A)} \begin{pmatrix}
d & -b \\
-c & a \\
\end{pmatrix}
$$

# Linear System of Equations

A system of linear equations can be represented in matrix form as $Ax=b$, where 
$A$ is a matrix of coefficients, $x$ is a column vector of variables, and $b$ is a column vector of results.

$$
Ax = b
$$

For example

$$
\begin{align*}
x + 2y &= 5 \\
3x + 4y &= 11
\end{align*}
$$

```{r}
eq1 <- function(x) { (5 - x) / 2 }
eq2 <- function(x) { (11 - 3*x) / 4 }

x_vals <- seq(-10, 10, length.out = 200)
y_vals1 <- eq1(x_vals)
y_vals2 <- eq2(x_vals)

plot(x_vals, y_vals1, type='l', col='blue', ylim=c(-10, 10), ylab='y', xlab='x')
lines(x_vals, y_vals2, col='red')
legend("topright", legend=c("x + 2y = 5", "3x + 4y = 11"), col=c("blue", "red"), lty=1)
```

```{r}
A <- matrix(c(1, 3, 2, 4), nrow=2)
b <- c(5, 11)

solution <- solve(A, b)

plot(x_vals, y_vals1, type='l', col='blue', ylim=c(-10, 10), ylab='y', xlab='x')
lines(x_vals, y_vals2, col='red')
points(solution[1], solution[2], pch=19, col='green')
legend("topright", legend=c("x + 2y = 5", "3x + 4y = 11", "Solution"), col=c("blue", "red", "green"), lty=1, pch=c(NA, NA, 19))
```

