---
title: "Game Theory Basics"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to Game Theory

Game theory is a field of mathematics that studies strategic interactions among rational decision-makers. It has applications in economics, political science, biology, computer science, and more.

### Key Concepts

1. **Players**: The decision-makers in the game.

2. **Strategies**: The plans of action available to players.

3. **Payoffs**: The outcomes or rewards players receive based on the strategies they choose.

---

In a typical game theory model, we have two players denoted as $A$ and $B$. Each player has a set of strategies available to them. Let's denote the strategies for player $A$ as $S_A = \{s_{A1}, s_{A2}, \ldots, s_{An}\}$ and for player $B$ as $S_B = \{s_{B1}, s_{B2}, \ldots, s_{Bm}\}$, where $n$ and $m$ are the number of strategies available to players $A$ and $B$, respectively.

The payoff functions for each player depend on the strategies chosen by both players and are represented as $u_A(s_{Ai}, s_{Bj})$ for player $A$ and $u_B(s_{Ai}, s_{Bj})$ for player $B$. Here, $u_A$ and $u_B$ denote the payoff functions for players $A$ and $B$, respectively, and $s_{Ai} \in S_A$, $s_{Bj} \in S_B$.

### Types of Games

1. **Cooperative vs Non-cooperative**: In cooperative games, players can form coalitions and make binding agreements. In non-cooperative games, binding agreements are not possible.

2. **Symmetric vs Asymmetric**: In symmetric games, the payoffs for playing a particular strategy depend only on the other strategies played, not on who is playing them. In asymmetric games, payoffs can differ for different players.

3. **Zero-sum vs Non-zero-sum**: In zero-sum games, one player's gain is another player's loss. In non-zero-sum games, it is possible for all players to gain or lose.


### Nash Equilibrium

A Nash Equilibrium is a situation in which no player can benefit by changing strategies while the other players keep theirs unchanged. In the Prisoner's Dilemma, the Nash Equilibrium is for both players to defect, even though both would be better off if they both cooperated.

A Nash Equilibrium is a condition where no player can gain by changing their strategy while the other player keeps theirs unchanged. Mathematically, a strategy pair $(s_{Ai}^*, s_{Bj}^*)$ is at Nash Equilibrium if:

$$
u_A(s_{Ai}^*, s_{Bj}^*) \geq u_A(s_{Ai}, s_{Bj}^*) , \forall s_{Ai} \in S_A
$$

$$
u_B(s_{Ai}^*, s_{Bj}^*) \geq u_B(s_{Ai}^*, s_{Bj}) , \forall s_{Bj} \in S_B
$$

This implies that given player $B$ is playing strategy $s_{Bj}^*$, player $A$ cannot improve their payoff by unilaterally changing their strategy from $s_{Ai}^*$ to any other strategy in $S_A$. Similarly, for player $B$, given player $A$ is playing strategy $s_{Ai}^*$.


## The Prisoner's Dilemma

One of the most famous examples in game theory is the Prisoner's Dilemma. Here, two individuals are arrested, but the police have insufficient evidence for a conviction. The police separate the prisoners and offer each the same deal: if one testifies against the other (defects) and the other remains silent (cooperates), the betrayer goes free, and the silent accomplice receives the full sentence. If both stay silent, both are sentenced to a short time in jail. If both betray, both receive a moderate sentence.

The payoff matrix can be represented as:

$$
\begin{array}{c|c|c}
& \text{Cooperate} & \text{Defect} \\
\hline
\text{Cooperate} & (R, R) & (S, T) \\
\hline
\text{Defect} & (T, S) & (P, P) \\
\end{array}
$$

Where:

- $T$ (Temptation) is the payoff for defecting against a cooperating player.
- $R$ (Reward) is the payoff for both cooperating.
- $P$ (Punishment) is the payoff for both defecting.
- $S$ (Sucker's payoff) is the payoff for cooperating against a defecting player.

Here, the Nash Equilibrium is $(D, D)$, where both players choose to defect. The only Nash Equilibrium is when both players choose to defect. Although cooperating might be better for both players collectively, it is not a Nash Equilibrium because if one player decides to cooperate, the best response for the other player is to defect.


```{r}
library(tidyverse)
payoff_matrix <- matrix(c(3, 0, 5, 1), nrow = 2, byrow = TRUE)
colnames(payoff_matrix) <- c("Cooperate", "Defect")
rownames(payoff_matrix) <- c("Cooperate", "Defect")
```

```{r}
simulate_game <- function(player1_decision, player2_decision) {
  payoff <- payoff_matrix[player1_decision, player2_decision]
  payoff[2] <- payoff_matrix[player2_decision, player1_decision]
  return(payoff)
}
```

```{r}
# Both Cooperate
both_cooperate <- simulate_game("Cooperate", "Cooperate")

# Both Defect
both_defect <- simulate_game("Defect", "Defect")

# Player 1 Cooperates, Player 2 Defects
one_cooperates <- simulate_game("Cooperate", "Defect")

# Store the results in a data frame for visualization
results <- data.frame(Scenario = c("Both Cooperate", "Both Defect", "One Cooperates"),
                      Payoff_Player1 = c(both_cooperate[1], both_defect[1], one_cooperates[1]),
                      Payoff_Player2 = c(both_cooperate[2], both_defect[2], one_cooperates[2]))
```

```{r}
results %>% 
  pivot_longer(cols = starts_with('Payoff'), 
               names_to = 'player',
               values_to = 'payoff') %>% 
  mutate(player = str_replace(player, "Payoff_", "")) %>%
  ggplot(aes(x = Scenario, y = payoff, fill = player)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Prisoner's Dilemma Payoffs", x = "Scenario", y = "Payoff",
       fill = "Player") +
  theme_minimal()
```

### With Learning